---
title: "Reproducible Research: Peer Assessment 1"
output:
  html_document:
    keep_md: true
---

# Reproducible Research: Peer Assessment 1

## Loading and preprocessing the data

We load data using `read.csv()` and convert the date column into dates.

```{r loaddata, echo=TRUE}
activity_data <- read.csv(unz('activity.zip', 'activity.csv'),
                          stringsAsFactor = FALSE)
# Convert Date to date types
activity_data[, 2] = as.Date(activity_data[, 2])
```

## What is mean total number of steps taken per day?

### Histogram of the total number of steps taken each day

```{r histogram, echo=TRUE}
steps_by_days <- aggregate(steps ~ date, data = activity_data, sum,
                           na.rm = TRUE)

library(ggplot2)
qplot(steps, data = steps_by_days, geom = 'histogram',
    xlab = "Total Steps", ylab = "Number of days",
    main = "Total steps per day - Histogram", binwidth = 1000)
```

### Mean and median of total number of steps taken per day

```{r meanmedian, echo = TRUE}
step_mean <- mean(steps_by_days$steps)
step_median <- median(steps_by_days$steps)
```
* The mean number of steps taken per day = **`r step_mean`**
* The median number of steps taken per day = **`r step_median`**


## What is the average daily activity pattern?

### Time series plot of 5-minute interval and average number of steps taken

```{r timeseries, echo = TRUE}
steps_by_interval <- aggregate(steps ~ interval,
                               data = activity_data,
                               mean, na.rm = TRUE)
qplot(data = steps_by_interval, x = interval, y = steps,
    ylab = "Average steps", main = "Time series plot") + geom_line()
```

### 5-minute interval containing the maximum number of steps

```{r maxInterval, echo = TRUE}
max_interval <- steps_by_interval[which.max(steps_by_interval$steps),]$interval
```
The 5 minute interval that contains the maximum number of steps is
**`r max_interval`**.

## Imputing missing values

### Total number of missing values in the dataset

```{r missing_values, echo = TRUE}
missing_vals <- sum(is.na(activity_data))
```
There are **`r missing_vals` missing values** in the dataset.

### Strategy for filling in missing values

We could use the strategy of filling in **missing values with mean** for that
 interval. The function to get steps given interval is as follows:

```{r missing_strategy, echo = TRUE}
intervalToSteps <- function(interval) {
    steps_by_interval[steps_by_interval$interval == interval, ]$steps
}
```

### New dataset with missing values filled in

We create a new dataset with missing values filled in.

```{r clean_dataset, echo = TRUE}
# Make a copy of dataset
filled_activity_data <- activity_data
for (i in 1:nrow(filled_activity_data)) {
    if (is.na(filled_activity_data[i, ]$steps)) {
        filled_activity_data[i, ]$steps <-
            intervalToSteps(filled_activity_data[i, ]$interval)
    }
}
```

### Analysis on Imputing

#### Histogram with clean data

```{r clean_histogram, echo=TRUE}
filled_steps_by_days <- aggregate(steps ~ date, data = filled_activity_data,
                                  sum, na.rm = TRUE)

qplot(steps, data = filled_steps_by_days, geom = 'histogram',
    xlab = "Total Steps", ylab = "Number of days",
    main = "Total steps per day - Histogram (filled data)", binwidth = 1000)
```

#### Mean and median of total number of steps taken per day, in clean data

```{r filled_meanmedian, echo = TRUE}
filled_step_mean <- mean(filled_steps_by_days$steps)
filled_step_median <- median(filled_steps_by_days$steps)
```
* The mean number of steps taken per day, in the filled dataset
= **`r filled_step_mean`**, where as the mean was **`r step_mean`** in original data
* The median number of steps taken per day, in the filled dataset
= **`r filled_step_median`**, where as the median was
**`r step_median`** in original data

## Are there differences in activity patterns between weekdays and weekends?

### Weekday factor in filled data

```{r weekday, echo = TRUE}
filled_activity_data$day <- ifelse(
    as.POSIXlt(filled_activity_data$date)$wday%%6 == 0, 'weekend', 'weekday')
filled_activity_data$day <- factor(filled_activity_data$day,
                                  levels = c('weekend', 'weekday'))
```

### Panel plot with time series plot comparing weekends and weekdays

```{r panel_plot, echo = TRUE, fig.width = 16}
library(lattice)
steps_by_interval_days <- aggregate(steps ~ interval + day,
                               data = filled_activity_data,
                               mean, na.rm = TRUE)

# To get the plot as displayed in Coursera
# xyplot(steps ~ interval | factor(day), data = steps_by_interval_days,
#       aspect = 1/2, type = "l")

qplot(data = steps_by_interval_days, x = interval, y = steps, facets = ~ day,
    ylab = "Average steps", main = "Time series plot") + geom_line()
```
